\section{Taxonomy of Persona Generation Methods}
Persona-generation methods cannot be understood from a single list of persona
types. Labels such as demographic, psychometric, cultural, character,
stakeholder, or individualized persona describe what attributes or identities
appear in a persona. They do not say where the persona information came from,
how the persona distribution was constructed, or what validity claim the persona
set is meant to support. We therefore classify persona-generation methods along
two orthogonal axes: construction technique and objective. The construction
axis asks where the persona information comes from before LLM enactment. The
objective axis asks what the persona set is meant to be valid for.

\subsection{Construction-Technique Axis}

Along the construction-technique axis, persona-generation methods differ by the
source used to construct persona-conditioning inputs: Authored Archetypes,
Model-Generated Personas, Population-Sampled Personas, and Trace-Grounded
Personas. This axis is determined by the anchor source, not by the final format.
Nearly every method can end in an LLM prompt, a natural-language profile, or an
agent memory. The relevant question is what source carries the substantive
persona information before enactment.

\begin{table}[t]
\centering
\small
\begin{tabular}{p{0.18\linewidth}p{0.22\linewidth}p{0.25\linewidth}p{0.24\linewidth}}
\toprule
\textbf{Family} & \textbf{Source} & \textbf{Construction logic} & \textbf{Main risk} \\
\midrule
Authored Archetypes &
Human designers, domain experts, stakeholders, or user researchers &
Manually craft a small number of interpretable representative types &
Subjectivity, stereotyping, overgeneralization, and weak representativeness \\
\addlinespace
Model-Generated Personas &
LLM priors plus broad prompts, labels, or weak constraints &
Ask the model to invent personas, attributes, backstories, or diverse sets &
Opaque model priors, prompt sensitivity, stereotypes, and incoherent joints \\
\addlinespace
Population-Sampled Personas &
Census data, surveys, panels, synthetic populations, or fused datasets &
Draw, weight, impute, calibrate, or synthesize anonymous records from target population structure &
Marginal-only alignment, missing variables, unobserved dependencies, and fusion assumptions \\
\addlinespace
Trace-Grounded Personas &
Interviews, logs, social media, profiles, histories, protocols, or other person-level evidence &
Reconstruct persona states from evidence about specific people or traces &
Privacy, consent, selection bias, memorization, and weak representativeness \\
\bottomrule
\end{tabular}
\caption{Construction-technique axis for persona generation.}
\label{tab:construction-axis}
\end{table}

\textbf{Authored Archetypes} are human-designed persona types. They inherit from
the HCI, requirements, product, and design tradition in which personas support
human reasoning, communication, and decision-making. In LLM-agent work, authored
or verified customer-persona artifacts appear in systems that bring customer
segments to life or refine personas with expert critique
\citep{arxiv250517156,arxiv260609637}. Their strength is interpretability and
domain relevance. Their weakness is that the type system may reflect the
designer's assumptions more than a target population.

\textbf{Model-Generated Personas} are invented primarily by an LLM from broad
prompts, labels, or weak constraints. They are attractive because they scale
quickly and produce vivid, varied surface descriptions. They are also the
clearest case where narrative fluency can be mistaken for methodological
grounding. Audits of open-ended persona brainstorming and critiques of
LLM-generated personas show that apparent diversity can hide stereotypes,
unstable sampling behavior, and invalid attribute combinations
\citep{arxiv250316527,arxiv260200044}. The model prior is doing much of the
sampling work, but that prior is usually opaque.

\textbf{Population-Sampled Personas} are top-down draws from population
structure. The anchor may be a census, survey, panel, synthetic-population
model, official statistic, or fused dataset. Their central claim is not merely
that each generated row sounds plausible, but that the persona set preserves
relevant marginal and joint structure. Examples include census-derived testbeds,
survey-derived persona prompts, and large population-persona datasets
\citep{arxiv240717387,arxiv251121722}.
% Citation TODO: add/verify a BibTeX key for NVIDIA Nemotron-Personas-USA.
% URL: https://huggingface.co/datasets/nvidia/Nemotron-Personas-USA
For this family, an eight-year-old persona with college education is not a
fluency failure; it is a joint-validity failure. Population-sampled methods are
therefore strongest when they report not only one-dimensional marginals, but
also dependencies such as age by education by household composition by region.

\textbf{Trace-Grounded Personas} are bottom-up reconstructions from
person-level evidence. The anchor may be an interview, longitudinal record,
social-media history, interaction log, profile, survey protocol, or design
research artifact. Such methods are often appropriate for individual fidelity,
behavioral calibration, and grounded user simulation. Examples include
social-media-grounded virtual populations, sociopsychological user-simulation
protocols, and digital-twin benchmarks \citep{arxiv250714922,arxiv260107110,arxiv251025536}.
Their risks differ from population-sampled methods: trace-grounded personas can
be richly evidenced and still be unrepresentative, privacy-sensitive, or
dependent on traces that were collected under selective conditions.

The same surface persona can therefore belong to different construction
families. ``A 45-year-old rural woman'' is an Authored Archetype if a designer
selected it, Model-Generated if an LLM invented it from a broad request for
diverse users, Population-Sampled if it was drawn from census or survey
structure, and Trace-Grounded if it summarizes a real participant or user.
Hybrids should be classified by their anchor source. A census scaffold enriched
with personality traits remains Population-Sampled if population structure
carries the validity claim. Social-media traces reweighted to population targets
remain Trace-Grounded with population calibration if person-level evidence is
the starting point.

\subsection{Objective Axis}

Along the objective axis, persona-generation methods differ by the claim their
persona sets are meant to support: population representation, coverage and
stress testing, individual fidelity, behavioral calibration, design
communication, bias and harm auditing, or agent/model evaluation. Objective
labels are multi-label. A single method can aim to represent a population,
calibrate behavior, and audit subgroup harms. Conversely, the same construction
family can serve different objectives.

\begin{table}[t]
\centering
\small
\begin{tabular}{p{0.22\linewidth}p{0.34\linewidth}p{0.32\linewidth}}
\toprule
\textbf{Objective} & \textbf{Question answered} & \textbf{Failure if confused} \\
\midrule
Population Representation &
Does the set represent a target population, including relevant marginals and joints? &
Treating vivid examples or balanced marginals as population validity \\
\addlinespace
Coverage and Stress Testing &
Does the set span rare, risky, difficult, or adversarial cases? &
Mistaking deliberate over-sampling of edge cases for prevalence estimates \\
\addlinespace
Individual Fidelity &
Does the persona preserve a specific person, trace, style, memory, or history? &
Treating a person-specific reconstruction as representative of a group \\
\addlinespace
Behavioral Calibration &
Do persona-conditioned agents reproduce observed responses, actions, or interactions? &
Treating response matching as proof that persona construction was valid \\
\addlinespace
Design Communication &
Do personas help humans reason, design, deliberate, or align stakeholders? &
Treating communicative usefulness as statistical representativeness \\
\addlinespace
Bias and Harm Auditing &
Does the method expose stereotypes, erasure, unfairness, or representational harm? &
Treating harm discovery as population simulation, or ignoring intersectional failures \\
\addlinespace
Agent / Model Evaluation &
Does persona conditioning test model behavior, role following, safety, alignment, or capability shifts? &
Treating model sensitivity to personas as evidence that the personas are valid \\
\bottomrule
\end{tabular}
\caption{Objective axis for persona generation.}
\label{tab:objective-axis}
\end{table}

\textbf{Population Representation} asks whether a persona set represents a
target population. This includes density matching, subgroup representation,
weighting, quota or stratified balance, and preservation of joint structure.
Survey-derived persona prompts and population-persona datasets are natural
examples \citep{arxiv251121722}.
% Citation TODO: add/verify a BibTeX key for NVIDIA Nemotron-Personas-USA.
% URL: https://huggingface.co/datasets/nvidia/Nemotron-Personas-USA
The relevant evidence is comparison to census, survey, panel, or official
reference data, with attention to both marginal and joint validity.

\textbf{Coverage and Stress Testing} asks whether the set spans important,
rare, difficult, or adversarial cases. Unlike population representation, the
goal may be to overrepresent low-prevalence situations because they reveal
failure modes. Persona generators that optimize support coverage rather than
density matching are a clear example \citep{arxiv260203545}. Evaluation should
therefore use coverage metrics, discovered failures, or robustness evidence
rather than prevalence accuracy alone.

\textbf{Individual Fidelity} asks whether a persona preserves a specific person
or trace. This objective is central to digital twins, longitudinal user
simulation, memory continuity, communication-style preservation, and
person-indexed modeling. Trace-grounded systems and digital-twin benchmarks
often combine individual fidelity with behavioral calibration
\citep{arxiv260107110,arxiv251025536}. The relevant evidence is person-level
comparison, not population-level representativeness.

\textbf{Behavioral Calibration} asks whether persona-conditioned agents
reproduce observed actions, responses, or interaction patterns. It can be
individual or aggregate, and it often appears in survey simulation,
recommendation, dialogue, and social-simulation settings
\citep{arxiv250714922,arxiv260107110,arxiv251121722}. Behavioral calibration
must be kept separate from persona validity. A coherent persona can be enacted
poorly, and a poorly grounded persona can sometimes produce plausible aggregate
behavior through model priors.

\textbf{Design Communication} asks whether personas help humans reason, design,
deliberate, or coordinate. This objective connects most directly to HCI and
requirements traditions, where personas are judged partly by interpretability,
stakeholder usefulness, and design relevance. Voice-of-customer-grounded
interviewable personas exemplify this objective \citep{arxiv260122288}.
Narrative enrichment is often important here, but narrative enrichment is a
means rather than a top-level objective.

\textbf{Bias and Harm Auditing} asks whether persona generation or persona
conditioning exposes stereotypes, erasure, unfairness, over-accentuation, or
other representational harms. Ethical audits of human and AI-crafted personas
and large-scale audits of open-ended persona brainstorming are central examples
\citep{arxiv250507850,arxiv260200044}. This objective often requires subgroup
and intersectional analysis, because harms can appear in attribute combinations
that are invisible in one-dimensional checks.

\textbf{Agent / Model Evaluation} asks what persona conditioning reveals about
LLM behavior. Many papers use personas not to create a population-valid persona
set, but to test role following, persona consistency, safety, alignment,
capability shifts, or vulnerability. Studies of persona collapse and
psychometric evaluation under persona conditioning fit this objective
\citep{arxiv260424698,arxiv251022170}. This category is included because
persona use in LLM research often evaluates the model rather than the persona
generation method.

\subsection{How the Axes Interact}

A construction family does not imply a validity claim. The same source can
serve different objectives, and the same objective can be pursued with different
sources. Model-generated personas can be used for bias and harm auditing, as in
open-ended persona-brainstorm audits \citep{arxiv260200044}, or for coverage and
stress testing, as in support-coverage-oriented persona generators
\citep{arxiv260203545}. Population-sampled personas can support population
representation and behavioral calibration when survey-derived prompts are used
to compare simulated and observed response distributions \citep{arxiv251121722}.
Trace-grounded personas can support individual fidelity and behavioral
calibration when they reconstruct participants, users, or digital twins from
person-level evidence \citep{arxiv260107110,arxiv251025536}.

This interaction is why the two slogans matter. Narrative vividness is not
population validity: a model-generated profile can be rich, memorable, and
useful for ideation without representing any target population. Persona validity
is not behavioral fidelity: a population-sampled or trace-grounded persona can
be well constructed while the LLM ignores, flattens, or overstates the persona
during enactment. Conversely, behavioral agreement can arise from shortcuts that
do not validate the persona set itself.

\subsection{Implications for Evaluation}

The taxonomy implies that evaluation should follow the claim. Population
Representation requires marginal and joint checks against a target reference.
Coverage and Stress Testing requires coverage metrics, scenario diversity, or
failure-discovery evidence. Individual Fidelity requires person-level
comparisons against traces, histories, or human judgments. Behavioral
Calibration requires benchmarks against observed responses, actions, or
interaction patterns. Design Communication requires evidence of usefulness,
interpretability, and design relevance for human stakeholders. Bias and Harm
Auditing requires subgroup, intersectional, and representational-harm analysis.
Agent / Model Evaluation requires role-following, consistency, safety,
alignment, or capability tests under persona conditioning. Later sections build
on this taxonomy to discuss persona-space design, sampling and identification,
evaluation standards, and ethical constraints.
